---
title: "Cyclistic's Bike-share Analysis"
author: "Oluwadunni"
date: "2024-09-11"
output: html_document
---


### Setting up my environment

Setting my R environment by loading the `tidyverse` and `conflicted` packages:

```{r load tidyverse, echo=TRUE, message=FALSE, warning=FALSE}
library(tidyverse)

```

```{r load conflicted}
library(conflicted)
```

**Managing conflicted packages**

Set `dplyr::filter` and `dplyr::lag` as the default choices

```{r manage conflicts, echo=TRUE, message=FALSE, warning=FALSE, paged.print=TRUE}
conflict_prefer("filter", "dplyr")
conflict_prefer("lag", "dplyr")
```

### Import Divy datasets
Notes: Importing dataframe for each quarter from `January2019` - `March 2020`
```{r import dataset, message=FALSE, warning=FALSE}
q1_2019 <- read_csv("Divvy_Trips_2019_Q1 - Divvy_Trips_2019_Q1.csv")
q2_2019 <- read_csv("Divvy_Trips_2019_Q2.csv")
q3_2019 <- read_csv("Divvy_Trips_2019_Q3.csv")
q4_2019 <- read_csv("Divvy_Trips_2019_Q4.csv")
q1_2020 <- read_csv("Divvy_Trips_2020_Q1 - Divvy_Trips_2020_Q1.csv")

```

## Dataframe cleaning and merging into a single file 

**Observing column names**
```{r observing column names}
colnames(q1_2019)
colnames(q2_2019)
colnames(q3_2019)
colnames(q4_2019)
colnames(q1_2020)
```
**Renaming columns for consistency**

```{r rename colums, message=FALSE, warning=FALSE}
(q1_2019 <- rename(q1_2019
                   ,ride_id = trip_id
                   ,rideable_type = bikeid
                   ,started_at = start_time
                   ,ended_at = end_time
                   ,start_station_name = from_station_name
                   ,start_station_id = from_station_id
                   ,end_station_name = to_station_name
                   ,end_station_id = to_station_id
                   ,member_casual = usertype
))
(q2_2019 <- rename(q2_2019
                   ,ride_id = `01 - Rental Details Rental ID`
                   ,rideable_type = `01 - Rental Details Bike ID`
                   ,started_at = `01 - Rental Details Local Start Time`
                   ,ended_at = `01 - Rental Details Local End Time`
                   ,start_station_name = `03 - Rental Start Station Name`
                   ,end_station_name = `02 - Rental End Station Name`
                   ,end_station_id = `02 - Rental End Station ID`
                   ,member_casual = `User Type`
                   ,start_station_id = `03 - Rental Start Station ID`
))
(q3_2019 <- rename(q3_2019
                   ,ride_id = trip_id
                   ,started_at = start_time
                   ,ended_at = end_time
                   ,rideable_type = bikeid
                   ,start_station_id = from_station_id
                   ,start_station_name = from_station_name
                   ,end_station_name = to_station_name
                   ,member_casual = usertype
                   ,end_station_id = to_station_id
))
(q4_2019 <- rename(q4_2019
                   ,ride_id = trip_id
                   ,started_at = start_time
                   ,ended_at = end_time
                   ,rideable_type = bikeid
                   ,start_station_id = from_station_id
                   ,start_station_name = from_station_name
                   ,end_station_name = to_station_name
                   ,member_casual = usertype
                   ,end_station_id = to_station_id
))
```

**Inspecting dataframes for inconsistencies**

```{r inspect dataframes, message=FALSE, warning=FALSE}
str(q1_2019)
str(q2_2019)
str(q3_2019)
str(q4_2019)
str(q1_2020)
```

**Convert data types for consistency**

```{r convert data types}
q1_2019 <-  mutate(q1_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)
                   ,started_at = as.POSIXct(started_at, format = "%Y-%m-%d %H:%M:%S", tz = "UTC")
                   ,ended_at = as.POSIXct(ended_at, format = "%Y-%m-%d %H:%M:%S", tz = "UTC"))
q2_2019 <-  mutate(q2_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)
                   ,started_at = as.POSIXct(started_at, format = "%Y-%m-%d %H:%M:%S", tz = "UTC")
                   ,ended_at = as.POSIXct(ended_at, format = "%Y-%m-%d %H:%M:%S", tz = "UTC"))

q3_2019 <-  mutate(q3_2019, ride_id = as.character(ride_id)
                   ,rideable_type = as.character(rideable_type)
                   ,started_at = as.POSIXct(started_at, format = "%Y-%m-%d %H:%M:%S", tz = "UTC")
                   ,ended_at = as.POSIXct(ended_at, format = "%Y-%m-%d %H:%M:%S", tz = "UTC"))

q4_2019 <- mutate(q1_2019, ride_id = as.character(ride_id)
                  ,rideable_type = as.character(rideable_type)
                  ,started_at = as.POSIXct(started_at, format = "%Y-%m-%d %H:%M:%S", tz = "UTC")
                  ,ended_at = as.POSIXct(ended_at, format = "%Y-%m-%d %H:%M:%S", tz = "UTC"))
q1_2020 <- mutate(q1_2020, started_at= as.POSIXct(started_at, format = "%Y-%m-%d %H:%M:%S", tz = "UTC")
                  ,ended_at = as.POSIXct(ended_at, format = "%Y-%m-%d %H:%M:%S", tz = "UTC"))
```

**Merge individual quarter's dataframe into one single dataframe**

```{r cache=FALSE}
all_trips <- bind_rows(q1_2019, q2_2019, q3_2019, q4_2019, q1_2020)
```

**Inspect columns of new data frame**

```{r}
colnames(all_trips)
```

**Remove discontinued fields from data frame**

Notes: Remove fields that were dropped such as Gender, start_lng .....

```{r}
all_trips <- all_trips %>%  
  select(-c(start_lat, start_lng, end_lat, end_lng, birthyear, gender,  tripduration, `01 - Rental Details Duration In Seconds Uncapped`, `05 - Member Details Member Birthday Year`, `Member Gender`  ))
```

**Inspecting new frame that has been created**

```{r}
colnames(all_trips)
```
**Retrieve Dimension for new data frame** 
```{r}
dim(all_trips)
```
**Generate data set summary**
```{r}
summary(all_trips)
```
**Preview first and last rows of dataset**
```{r}
head(all_trips)
tail(all_trips)
```
**Distribution of membership types**
```{r}
table(all_trips$member_casual)

```

**Reassign desired values in `member_casual` column**

```{r reassign }
all_trips <-  all_trips %>% 
  mutate(member_casual = recode(member_casual
                                ,"Subscriber" = "member"
                                ,"Customer" = "casual"))

```

**Add columns to dataframe**

Note: Add columns that list the date, month, day, and year of each ride

```{r add columns}
all_trips$date <- as.Date(all_trips$started_at) 
all_trips$month <- format(as.Date(all_trips$date), "%m")
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")

```

**Add a calculated field to the dataframe**

Note: Add `ride length` to calculate the duration for each trip in seconds

```{r add ride length}
all_trips$ride_length <- difftime(all_trips$ended_at,all_trips$started_at)

```

**Remove bad data**

Note: Remove observations where `ride_length` was negative or where bikes were taken out for maintenance

```{r}
all_trips_v2 <- all_trips[!(all_trips$start_station_name == "HQ QR" | all_trips$ride_length<0),]

```

### Descriptive analysis on `ride_length`

Notes: All figures are in seconds

**Straight average of all trips(total ride length / rides)**

```{r Mean}
mean(all_trips_v2$ride_length) 

```

**Median (midpoint number in the ascending array of ride lengths)**

```{r Median}
median(all_trips_v2$ride_length)
```

**Longest ride**

```{r Max trip}
max(all_trips_v2$ride_length)
```

**Shortest ride**

```{r Min trip}
min(all_trips_v2$ride_length)
```

**Compare members and casual users**

Notes: Calculations for mean, median,max and min for each membertype 
```{r average rides grouped by member type}
all_trips_v2 %>% 
  group_by(member_casual) %>%
  summarize(mean_ride_length = mean(ride_length), median_ride_length = median(ride_length), min_ride_length = min(ride_length), max_ride_length = max(ride_length))
```
**Organize ``day_of_week` column**

```{r}
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))

```
**Average ride time by each day for members vs casual users**

```{r}
all_trips_v2 %>% 
  group_by(member_casual, day_of_week) %>% 
  summarise(mean_ride_length = mean(ride_length))
```

**Analyze ridership data by type and `day_of the week`**
```{r} 
all_trips_v2 %>%
  group_by(member_casual, day_of_week) %>% 
  summarise(number_of_rides = n()
            ,average_duration = mean(ride_length)) %>% 
  arrange(member_casual, day_of_week)		
  
```
## Visualization
**Visual to compare average ride_length for member type**
```{r visuals}
all_trips_v2 %>% 
  group_by(member_casual, day_of_week) %>% 
  summarise(mean_ride_length = mean(ride_length)) %>% 
  arrange(member_casual, day_of_week)  %>% 
  ggplot(aes(x = day_of_week, y = mean_ride_length, fill = member_casual)) +
  geom_col(position = "dodge")+
  labs(title = "Ride Length by Membership and Day", subtitle ="A Comparative Analysis of Ride Durations", caption = ("Data from January 2019 - March 2020"), x = "Day of week", y = "Average ride length(secs)", fill = "Membership Type")
  
```

